Practical Optional Types for Clojure

### 2015-02-28 formal-model.tex
paths

\paragraph{Operational semantics} We define the dynamic semantics for \lambdatc{}
in a big-step style using an environment, following~\citet{TF10}.
We include both errors and a \wrong{} value, which is provably ruled out by the
type system.
The main judgement is \opsem{\openv{}}{\e{}}{\definedreduction{}}
which states that \e{} evaluates to answer \definedreduction{} in environment
\openv{}.
A notable difference is \nil{} is a false value, which affects the
semantics of \ifliteral{}:

@@

We now present \HMapliteral{} types in detail.
\figref{main:figure:hmapsyntax} extends the syntax for terms and types.
The syntax 
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
we introduced in \secref{sec:overview}
is now written \HMapgeneric{\mandatory{}}{\absent{}},
where \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} and is {\partial{}} otherwise. 
The \mandatory{} entry specifies a set of known present keyword entries and \absent{} tracks known absent keys.
To ease presentation, 
if a HMap is declared as \complete{} then \absent{} implicitly contains all other keywords not specified in \mandatory{}.
%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
Keys cannot be both present and absent.

@@

\figref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
We only handle unions of HMaps in T-GetHMap, where every element in the union
contains the key we are looking up, in particular to support operations like
looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf}).
The object in each T-Get rule is now more complicated than in T-Local---the 
next section discusses this in detail.
T-Assoc extends a HMap with a mandatory entry.

@@

Recall the first insight of occurrence typing---not
only do we want to reason about components of the runtime environment, that
is, variables, but \emph{subcomponents} of the environment.
We briefly mentioned paths, written \path{\pathelem{}}{\x{}}, but we concentrated on
reasoning only about the variable \x{}---the notation \pathelem{} is a nested sequence of \emph{path elements}
applied right-to-left to a variable \x{}.
\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
to reason about selector operations on cons cells.
We instead want to reason about HMap lookups.

>>>>

\paragraph{Operational semantics} We define the dynamic semantics for \lambdatc{}
in a big-step style using an environment, following~\citet{TF10}.
We include both errors and a \wrong{} value, which is provably ruled out by the
type system.
The main judgement is \opsem{\openv{}}{\e{}}{\definedreduction{}}
which states that \e{} evaluates to answer \definedreduction{} in environment
\openv{}. We chose to omit the core rules (see \figref{appendix:figure:opsem})
however a notable difference is \nil{} is a false value, which affects the
semantics of \ifliteral{}:

@@

We now present \HMapliteral{} types in detail.
\figref{main:figure:hmapsyntax} extends the syntax for terms and types.

We write \HMapgeneric{\mandatory{}}{\absent{}}
where {\mandatory{}} is a map of \emph{present} entries (mapping keywords to types),
\absent{} is a set of keyword keys that are known to be \emph{absent}
and
\completenessmeta{} is {\complete{}} (``complete'') if the map is fully specified by \mandatory{},
and {\partial{}} (``partial'') if there are \emph{unknown} entries.
To ease presentation, 
if a HMap is declared as \complete{} then \absent{} implicitly contains all keywords not in the domain of \mandatory{}.
%\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. 
Keys cannot be both present and absent.

@@

\figref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
Lookups on unions of HMaps are only supported in T-GetHMap, 
in particular to support
looking up \clj{:op} on a map of type \clj{Expr} (\egref{example:decleaf})
where every element in the union
contains the key we are looking up.
The objects in the T-Get rules are more complicated than those in T-Local---the 
next section discusses this in detail.
Finally T-AssocHMap extends a HMap with a mandatory entry while preserving completeness
and absent entries, and enforcing ${\k{}} \not\in {\absent{}}$ to prevent badly
formed types.

@@

Recall the first insight of occurrence typing---we can reason
about specific \emph{parts} of the runtime environment
using propositions.
The way to refer to \emph{parts} of the runtime environment with
occurrencing is via \emph{path elements}.
A \emph{path} consists of a series of path elements
applied right-to-left to a variable
written
\path{\pathelem{}}{\x{}}.
\citet{TF10} introduce the path elements \carpe{} and \cdrpe{}
to reason about selector operations on cons cells.
We instead want to reason about HMap lookups and calls to \classconst{}.
